\centerline{\xiaoyi\bfseries Abstract}

Clinical decision-making is inherently longitudinal — clinicians must synthesize evidence across repeated hospitalizations, diagnostic tests, and evolving treatment responses spanning months or years. However, current benchmarks for medical large language model (LLM) predominantly evaluate single-encounter behavior, focusing on model knowledge, tool-use accuracy and fact retrieval rather than the cross-visit temporal integration that real clinical practice demands. To address this gap, this thesis introduces the Long-Horizon Reasoning Task of LongMedBench, a benchmark component constructed from real-world Electronic Health Record (EHR) data (MIMIC-IV) that evaluates whether LLM agents can reason over longitudinal clinical history before selecting clinical decisions or actions. We design a three-tier memory architecture operating at three granularities: Note Memory (visit-level summaries), Event Memory (fine-grained structured event sequences), and Contextual Memory (immediate visit context). Based on the datasets, we systematically compare memory augmentation strategies including naive long-context, retrieval-augmented generation (RAG), and agent memory systems (Mem0). Through experiments on GPT-5-mini, DeepSeek-v3.2, and Qwen-Turbo, we find that decision-making accuracy across all models remains modest, and critically, that memory augmentation architectures provide no significant improvement over the naive long-context baseline. Decision accuracy depends primarily on the immediate clinical context, with historical information contributing negligible marginal value regardless of memory granularity or retrieval strategy. This central finding reveals a retrieval-reasoning gap: current architectures can locate relevant historical facts but cannot effectively integrate them into proactive clinical reasoning. We discuss the implications of this gap for medical AI agent design and propose directions for memory architectures that go beyond retrieval to support structured longitudinal reasoning.

\vspace{2em}
\noindent\textbf{Keywords:} Medical AI Agents, Clinical Decision-Making, Electronic Health Records, Long-Horizon Reasoning, Retrieval-Augmented Generation, Agent Memory Systems
