\centerline{\xiaoyi\bf Appendix}
\addcontentsline{toc}{chapter}{Appendix}

This appendix reorganizes the supplementary material around the benchmark artifacts used to construct and evaluate LongMedBench. Code-like artifacts are shown in shaded blocks with a colored header. Long notes, reports, and model responses are shortened only for readability while preserving the original schema and decision content.

\section*{A1 Dataset Pipeline}
\addcontentsline{toc}{section}{A1 Dataset Pipeline}

The dataset pipeline starts from raw MIMIC-IV notes and structured tables, converts each admission into a typed clinical event stream, and then derives the current-visit interaction context used by the agentic decision task. A representative patient-level sequence record contains 813 ordered events for Patient~P000001, including demographics, admissions, labs, microbiology, imaging, medications, procedures, and discharges.

\subsection*{A1.1 Raw Discharge Note Example}
\addcontentsline{toc}{subsection}{A1.1 Raw Discharge Note Example}

The raw note table stores one discharge summary per row. A representative row has the schema \texttt{note\_id}, \texttt{subject\_id}, \texttt{hadm\_id}, \texttt{note\_type}, \texttt{note\_seq}, \texttt{charttime}, \texttt{storetime}, and \texttt{text}.

\begin{promptblock}{Raw Discharge Note Row}
note_id: 10000032-DS-21
subject_id: 10000032
hadm_id: 22595853
note_type: DS
charttime: 2180-05-07 00:00:00
storetime: 2180-05-09 15:26:00
text: Name: ___  Unit No: ___
Admission Date: ___  Discharge Date: ___
Service: MEDICINE
Allergies: No Known Allergies / Adverse Drug Reactions
Chief Complaint: Worsening ABD distension and pain
Major Surgical or Invasive Procedure: Paracentesis
History of Present Illness: ___ HCV cirrhosis c/b ascites,
hiv on ART, h/o IVDU, COPD, bipolar, PTSD, presented from OSH ED
with worsening abdominal distension and pain ...
\end{promptblock}

During preprocessing, admission-facing fields such as chief complaint and history of present illness initialize the current visit, while discharge-facing fields become visit-level Note Memory for later visits.

\subsection*{A1.2 Structured Admission and Discharge Events}
\addcontentsline{toc}{subsection}{A1.2 Structured Admission and Discharge Events}

In the sequenced patient file, a visit is represented by explicit boundary events. For example, the first retained visit of P000001 begins with an admission event and ends with a discharge event.

\begin{promptblock}{Structured Admission and Discharge Events}
{
  "event_id": "P000001-V1-adm",
  "event_type": "ADMISSION",
  "timestamp": "2120-08-06 20:15:00",
  "visit_ref": "V1",
  "content": {
    "location": "PACU",
    "admission_type": "EW EMER.",
    "chief_complaint": "hypotension and lethargy s/p RFA",
    "history_of_present_illness": "Mr. ___ is a ___ year old man ...",
    "insurance": "Medicaid"
  }
}

{
  "event_id": "P000001-V1-dis",
  "event_type": "DISCHARGE",
  "timestamp": "2120-08-14 19:30:00",
  "visit_ref": "V1",
  "content": {
    "discharge_time": "2120-08-14 19:30:00",
    "discharge_location": "HOME",
    "discharge_note": {...}
  }
}
\end{promptblock}

\subsection*{A1.3 Radiology and Imaging Event Example}
\addcontentsline{toc}{subsection}{A1.3 Radiology and Imaging Event Example}

Radiology reports are converted into imaging events with a timestamp and report text. The same patient sequence contains the following imaging event.

\begin{promptblock}{Radiology to Imaging Event}
{
  "event_id": "P000001-V1-E2",
  "event_type": "IMAGING",
  "timestamp": "2120-08-06 07:03:00",
  "visit_ref": "V1",
  "content": {
    "timestamp": "2120-08-06 07:03:00",
    "report": "CONTRAST-ENHANCED CT OF LIVER AND UPPER ABDOMEN\n\n"
              "INDICATION: ___ male with known cirrhosis and abnormalities "
              "in segment II and segment VI of the right lobe of liver. ..."
  }
}
\end{promptblock}

For Context Memory, this event is rendered as an assistant action such as \texttt{order\_imaging} followed by an environment observation containing the available radiology result.

\section*{A2 Memory Construction}
\addcontentsline{toc}{section}{A2 Memory Construction}

LongMedBench uses three memory forms. Note Memory stores previous-visit discharge summaries, Event Memory stores previous-visit structured event streams, and Context Memory stores the dialogue-style state of the current visit before a decision point.

\subsection*{A2.1 LLM Convert}
\addcontentsline{toc}{subsection}{A2.1 LLM Convert}

Context Memory is the only memory type that uses an LLM during construction. The code first maps structured events to semantic actions, then asks the LLM to write the assistant-side rationale field under strict grounding constraints.

\begin{promptblock}{ContextMemory LLM Conversion Prompt}
You write the 'reason' field for the next assistant action in a
doctor-agent dialogue.

Hard constraints:
- You may ONLY use information present in the provided prefix_messages.
- Do NOT invent new patient facts, findings, diagnoses, or outcomes.
- Do NOT mention results of the current action (results are not available yet).
- Keep it 1-2 sentences, not exceeding 50 words, workflow-rationale focused.

Output ONLY JSON: {"reason": string, "constraints_ok": boolean}.

User payload:
{
  "prefix_messages": <system, admission, and previous interaction messages>,
  "current_decision": {"action": <action>, "args": <arguments>}
}
\end{promptblock}

The remaining memory types are constructed automatically. Note Memory is produced by extracting the complete discharge-note text from previous visits. Event Memory is produced by chronological serialization of structured EHR events. Neither requires LLM rewriting.

\subsection*{A2.2 Memory Templates}
\addcontentsline{toc}{subsection}{A2.2 Memory Templates}

\begin{promptblock}{NoteMemory Template}
[Visit P000001-V1 NoteMemory]
Chief Complaint: hypotension and lethargy s/p RFA
History of Present Illness: Mr. ___ is a ___ year old man with
EtOH and HCV cirrhosis, c/b ascites, prior variceal bleeds, recent MRI
with 2 lesions suspicious for ___, who was admitted s/p RFA with
lethargy and hypotension. ...
Discharge Location: HOME
Discharge Summary: <complete discharge-note fields from raw note>
\end{promptblock}

\begin{promptblock}{EventMemory Template}
PROCEDURE  P000001-V1-E1  2120-08-06
  items: Closed percutaneous needle biopsy of liver;
         Percutaneous ablation of liver lesion or tissue

LAB        P000001-V1-E8  2120-08-08 05:10:00
  items: Red Blood Cells 3.29 m/uL abnormal; RDW 16.7% abnormal;
         Alanine Aminotransferase included in Chemistry panel

MICRO      P000001-V1-E11 2120-08-08 11:45:00
  specimen: BLOOD CULTURE; result: NO GROWTH

MEDICATION P000001-V1-E4  2120-08-06 21:00:00
  items: Morphine Sulfate IV 0.5-1 mg ordered;
         Ondansetron IV 4 mg ordered
\end{promptblock}

\begin{promptblock}{ContextMemory Template}
{
  "patient_id": "P000001",
  "visit_id": "P000001-V12",
  "messages": [
    {"role": "system", "content": "You are a doctor agent ..."},
    {"role": "user", "content": "Gender: M. Age at first visit: 48 ..."},
    {"role": "assistant", "content": {
      "reason": "The patient has atypical chest pain and bilateral lower
                 extremity edema, warranting a chest X-ray ...",
      "action": "order_imaging",
      "args": {"modality": "xray", "target": "chest",
               "timestamp": "2121-12-16 19:05:00"}
    }},
    {"role": "user", "content": {
      "name": "environment",
      "observation_type": "imaging_results",
      "results": [{"test": "chest xray", "content": "..."}]
    }}
  ]
}
\end{promptblock}

\section*{A3 Question Generation}
\addcontentsline{toc}{section}{A3 Question Generation}

The agentic decision task uses three question templates generated from the current event prefix and future event stream. The examples below are representative generated questions for Patient~P000001.

\begin{table}[!h]
    \centering
    \caption{Question templates for the three agentic decision subtasks.}
    \label{tab:appendix_question_templates}
    {\fontsize{8}{9.2}\selectfont
    \setlength{\tabcolsep}{3pt}
    \renewcommand{\arraystretch}{1.15}
    \begin{tabular*}{\textwidth}{p{0.15\textwidth}@{\extracolsep{\fill}}p{0.27\textwidth}p{0.31\textwidth}p{0.18\textwidth}}
        \toprule
        \textbf{Type} & \textbf{Template} & \textbf{Example options} & \textbf{Answer form} \\
        \midrule
        T3-N Next Action & Which of the next step is most appropriate? & \texttt{perform\_procedure}, \texttt{medication}, \texttt{ask\_question}, \texttt{order\_microbiology}, \texttt{order\_labs}, \texttt{order\_imaging} & Weighted action dictionary \\
        T3-A Argument & Which imaging order is most appropriate next? & \texttt{CT head}, \texttt{MRI brain}, \texttt{CT abdomen}, \texttt{XRAY abdomen}, \texttt{CT chest}, \texttt{US pelvis} & Weighted argument dictionary \\
        T3-D Discharge & Should the patient be discharged within 6h? & \texttt{Yes}, \texttt{No} & Weighted binary dictionary \\
        \bottomrule
    \end{tabular*}}
\end{table}

\begin{promptblock}{Expected Model Output}
{"reason": "The patient has abnormal lab results...", "answer": ["order_labs"]}
\end{promptblock}

For T3-A questions, candidate answers are collected from future events in a 24-hour window after the current timestamp and weighted by temporal decay. T3-D instead compares the current timestamp with the discharge time using a six-hour horizon.

\section*{A4 Agent Configuration}
\addcontentsline{toc}{section}{A4 Agent Configuration}

Each evaluation condition is defined by two axes: the historical memory source, either Note Memory or Event Memory, and the delivery strategy, either long-context prompting, retrieval-augmented generation (RAG), or Mem0-style managed memory. The current-visit Context Memory is always included as the immediate interaction state.

\begin{table}[!h]
    \centering
    \caption{Agent context structure under each memory-delivery setting.}
    \label{tab:appendix_agent_config}
    {\fontsize{8}{9.2}\selectfont
    \setlength{\tabcolsep}{3pt}
    \renewcommand{\arraystretch}{1.15}
    \begin{tabular*}{\textwidth}{p{0.18\textwidth}@{\extracolsep{\fill}}p{0.22\textwidth}p{0.50\textwidth}}
        \toprule
        \textbf{Setting} & \textbf{Historical source} & \textbf{Context structure} \\
        \midrule
        Note + context & Previous discharge notes & System prompt, current-visit context messages, packed previous-visit discharge notes, user question. \\
        Event + context & Previous event streams & System prompt, current-visit context messages, packed chronological events from previous visits, user question. \\
        Note + RAG & Previous discharge notes indexed for retrieval & System prompt and current context, retrieved note snippets relevant to the question, user question. \\
        Event + RAG & Previous event streams indexed for retrieval & System prompt and current context, retrieved event snippets relevant to the question, user question. \\
        Note + Mem0 & Previous discharge notes stored as managed memories & System prompt and current context, Mem0-recalled note memories, user question. \\
        Event + Mem0 & Previous event streams stored as managed memories & System prompt and current context, Mem0-recalled event memories, user question. \\
        \bottomrule
    \end{tabular*}}
\end{table}

\section*{A5 Case Study}
\addcontentsline{toc}{section}{A5 Case Study}

This section provides concrete qualitative cases that explain why aggregate accuracy changes are small even when memory configuration changes individual decisions. Each case is centered on the original interaction-style JSON. Explanatory text is included only to summarize the behavioral pattern illustrated by the JSON.

A read-only inspection of the analysis artifacts showed that event and note memories often change model outputs even when the overall benchmark average remains close. In the shared 2,202-question overlap, event--note prediction changes occurred in 662 questions for the long-context agent (30.06\%), 807 questions for Mem0 (36.65\%), and 740 questions for RAG (33.61\%). The cases below therefore focus on how those changes arise rather than treating the mean score as the only evidence.

\paragraph{Case 1: Retrieval helps when local evidence reinforces the current trajectory.}
\begin{promptblock}{Original Interaction JSON: P000001-V12-T0002-T3-N-0}
{
  "qid": "P000001-V12-T0002-T3-N-0",
  "qtype": "T3-N",
  "timestamp": "2121-12-16 19:10:00",
  "question": "Which of the next step is most appropriate?",
  "options": [
    "perform_procedure",
    "medication",
    "ask_question",
    "order_microbiology",
    "order_labs",
    "order_imaging"
  ],
  "answer": {"order_imaging": 1.0},
  "context_messages": [
    {
      "role": "system",
      "content": "You are a doctor agent. You can perform these semantic actions: ask_question, order_labs, order_microbiology, order_imaging, perform_procedure, medication, discharge. The environment returns only information that exists in this admission record; otherwise it returns not_available. Do not invent new findings."
    },
    {
      "role": "user",
      "content": "Gender: M. Age at first visit: 48. Chief complaint: Chest Pa. HPI: male with HCV and EtOH cirrhosis on the transplant list, complicated by HCC s/p TACE/RFA and ascites, recurrent C. difficile colitis, increasing abdominal distention, bilateral lower extremity swelling, and occasional non-pleuritic, non-exertional chest pain for 3 days."
    },
    {
      "role": "assistant",
      "content": {
        "reason": "Atypical chest pain and bilateral lower extremity edema warrant chest X-ray evaluation.",
        "action": "order_imaging",
        "args": {"modality": "xray", "target": "chest", "timestamp": "2121-12-16 19:05:00"}
      }
    },
    {"role": "environment", "content": "chest xray -> Limited, negative."},
    {
      "role": "assistant",
      "content": {
        "reason": "Bilateral lower extremity edema warrants Doppler ultrasound to evaluate for deep vein thrombosis.",
        "action": "order_imaging",
        "args": {"modality": "doppler", "target": "free text body part", "timestamp": "2121-12-16 19:10:00"}
      }
    },
    {"role": "environment", "content": "free text body part doppler -> Addendum: A limited Color Doppler and Spectral Analysis Study was also performed."}
  ],
  "predictions": {
    "note_context": {"answer": ["order_labs"], "score": 0.0},
    "event_context": {"answer": ["order_labs"], "score": 0.0},
    "note_rag": {"answer": ["order_imaging"], "score": 1.0},
    "event_rag": {"answer": ["order_imaging"], "score": 1.0},
    "note_mem0": {"answer": ["order_labs"], "score": 0.0},
    "event_mem0": {"answer": ["order_labs"], "score": 0.0}
  }
}
\end{promptblock}

RAG succeeds because the retrieved material reinforces the current imaging trajectory, while Mem0 retrieves clinically plausible but temporally mismatched memories that answer a different situation.

\paragraph{Case 2: More memory is not always better.}
\begin{promptblock}{Original Interaction JSON: P000003-V16-T0009-T3-N-0}
{
  "memory_type": "event",
  "qid": "P000003-V16-T0009-T3-N-0",
  "qtype": "T3-N",
  "question": "Which of the next step is most appropriate?",
  "gt": ["medication"],
  "baseline": ["order_labs"],
  "baseline_reason": "The patient is undergoing R-ICE chemotherapy and has abnormal lab results, indicating the need for further laboratory testing to monitor for potential complications.",
  "preds": {
    "ctx0": ["medication"],
    "ctx2": ["medication"],
    "ctx5": ["order_labs"]
  },
  "reasons": {
    "ctx0": "The patient has abnormal lab results and is receiving chemotherapy, so ordering furosemide is necessary to manage potential fluid overload or related complications.",
    "ctx2": "The patient has abnormal lab results and is receiving chemotherapy, so ordering furosemide is necessary to manage potential fluid overload or related complications.",
    "ctx5": "The patient is undergoing R-ICE chemotherapy and has abnormal lab results, indicating the need for further laboratory testing to monitor for potential complications."
  }
}
\end{promptblock}

The zero-visit and two-visit contexts choose the correct action, but the five-visit context shifts back to laboratory monitoring. This is a non-monotonic context-length effect: adding history can dilute the current action signal rather than clarify it.

\paragraph{Case 3: Event memory can preserve action-relevant signals.}
\begin{promptblock}{Original Interaction JSON: P000002-V15-T0031-T3-N-0}
{
  "agent": "agentic_decision",
  "qid": "P000002-V15-T0031-T3-N-0",
  "qtype": "T3-N",
  "question": "Which of the next step is most appropriate?",
  "event_pred": ["medication"],
  "note_pred": ["order_labs"],
  "event_reason": "The patient has abnormal lab results and a positive respiratory culture, requiring targeted antibiotic therapy based on susceptibility results. Therefore, the most appropriate next step is to administer medication.",
  "note_reason": "The patient has multiple abnormal lab results and a positive respiratory culture, indicating the need for further diagnostic evaluation and targeted treatment. Ordering labs will help guide management.",
  "gt": ["medication"]
}
\end{promptblock}

Both memories recognize abnormal labs and a positive culture, but event memory converts that evidence into the next intervention whereas note memory remains at diagnostic evaluation.

\paragraph{Case 4: Retrieved memory can override the current clinical target.}
\begin{promptblock}{Original Interaction JSON: P000001-V12-T0002-T3-A-1}
{
  "memory_type": "note",
  "qid": "P000001-V12-T0002-T3-A-1",
  "qtype": "T3-A",
  "question": "Which imaging order is most appropriate next?",
  "gt": ["XRAY abdomen"],
  "baseline": ["XRAY abdomen"],
  "baseline_reason": "The patient has bilateral lower extremity edema, which is more consistent with a systemic condition such as liver disease or heart failure rather than a localized issue in the extremities. Therefore, an imaging study of the abdomen would be more appropriate to evaluate for underlying causes such as ascites or liver pathology.",
  "preds": {
    "llm_agent": ["XRAY abdomen"],
    "mem0_agent": ["CT head"],
    "rag_agent": ["CT chest"]
  },
  "reasons": {
    "llm_agent": "The patient has bilateral lower extremity edema, which is more consistent with a systemic condition such as liver disease or heart failure rather than a localized issue in the extremities. Therefore, an imaging study of the abdomen would be more appropriate to evaluate for underlying causes such as ascites or liver pathology.",
    "mem0_agent": "Patient has a reported headache during this admission and a noncontrast CT head is the most appropriate initial, rapid study to evaluate acute intracranial causes. Other abdominal and chest imaging have recently been performed and are less relevant to the new symptom.",
    "rag_agent": "The patient has chest pain with a limited/negative chest x-ray and a nondiagnostic/limited lower-extremity Doppler; the next appropriate study to evaluate for pulmonary embolism or other intrathoracic causes is a CT chest."
  }
}
\end{promptblock}

Mem0 and RAG retrieve recognizable clinical signals, but those signals redirect the answer away from the abdominal ascites and liver-pathology trajectory that supports the ground-truth order.

\paragraph{Case 5: Retrieval can correct an immediate-context error.}
\begin{promptblock}{Original Interaction JSON: P000001-V20-T0002-T3-N-0}
{
  "memory_type": "note",
  "qid": "P000001-V20-T0002-T3-N-0",
  "qtype": "T3-N",
  "question": "Which of the next step is most appropriate?",
  "gt": ["order_labs"],
  "baseline": ["order_imaging"],
  "baseline_reason": "The patient has a history of HCC and cirrhosis, and the current presentation includes right upper quadrant pain and abdominal distention. An ultrasound of the right upper quadrant is needed to evaluate for possible complications.",
  "preds": {
    "llm_agent": ["order_imaging"],
    "mem0_agent": ["order_labs"],
    "rag_agent": ["order_labs"]
  },
  "reasons": {
    "llm_agent": "The patient has a history of HCC and cirrhosis, and the current presentation includes right upper quadrant pain and abdominal distention. An ultrasound of the right upper quadrant is needed to evaluate for possible complications.",
    "mem0_agent": "RUQ ultrasound excluded biliary obstruction but the patient has cirrhosis, ascites, and was being evaluated for infection; laboratory studies are needed to assess for leukocytosis, renal/hepatic function, and guide further management.",
    "rag_agent": "Right upper quadrant ultrasound excluded biliary obstruction/choledocholithiasis. The patient with cirrhosis, splenomegaly, and perihepatic ascites needs laboratory evaluation to assess for infection, hepatic decompensation, and guide need for paracentesis or other interventions."
  }
}
\end{promptblock}

This positive case shows that memory can help when it supplies a completed prior diagnostic step that changes the next action.

\paragraph{Case 6: Note memory can over-emphasize narrative diagnosis.}
\begin{promptblock}{Original Interaction JSON: P000007-V11-T0033-T3-N-0}
{
  "agent": "mem0_agent",
  "qid": "P000007-V11-T0033-T3-N-0",
  "qtype": "T3-N",
  "question": "Which of the next step is most appropriate?",
  "event_pred": ["order_labs"],
  "note_pred": ["medication"],
  "event_reason": "The patient has persistent severe electrolyte abnormalities (hypokalemia, hypocalcemia, hypomagnesemia), metabolic acidosis and rising creatinine despite ongoing replacement; repeating labs/ABG and electrolytes is the next appropriate step to assess response and guide further therapy.",
  "note_reason": "The patient has a stool culture positive for Salmonella species and ongoing severe electrolyte abnormalities requiring continued IV/PO repletion and initiation/adjustment of targeted antimicrobial therapy in an immunocompromised host. Thus the immediate next step is medication management.",
  "baseline": {
    "event": ["order_labs"],
    "note": ["order_labs"]
  },
  "gt": ["order_labs"],
  "event_hit_gt": true,
  "note_hit_gt": false
}
\end{promptblock}

The note-memory answer foregrounds a high-level infectious diagnosis, whereas the event-memory answer preserves the immediate monitoring need caused by severe electrolyte and acid-base abnormalities.

\begin{table}[!h]
    \centering
    \caption{Qualitative patterns illustrated by the A5 cases.}
    \label{tab:appendix_case_patterns}
    {\fontsize{7.2}{8.4}\selectfont
    \setlength{\tabcolsep}{3pt}
    \renewcommand{\arraystretch}{1.16}
    \begin{tabular*}{\textwidth}{p{0.22\textwidth}@{\extracolsep{\fill}}p{0.18\textwidth}p{0.22\textwidth}p{0.30\textwidth}}
        \toprule
        \textbf{Case} & \textbf{Ground truth} & \textbf{Main contrast} & \textbf{Interpretation} \\
        \midrule
        \texttt{P000001-V12-T0002-T3-N-0} & \texttt{order\_imaging} & RAG correct; context and Mem0 incorrect & Retrieval helps when it reinforces the current imaging trajectory. \\
        \texttt{P000003-V16-T0009-T3-N-0} & \texttt{medication} & ctx0/ctx2 correct; ctx5 incorrect & Longer history can dilute the action signal. \\
        \texttt{P000002-V15-T0031-T3-N-0} & \texttt{medication} & event correct; note incorrect & Event memory can preserve action-level evidence better than narrative notes. \\
        \texttt{P000001-V12-T0002-T3-A-1} & \texttt{XRAY abdomen} & baseline correct; Mem0/RAG incorrect & Retrieved memories can be plausible but target the wrong clinical thread. \\
        \texttt{P000001-V20-T0002-T3-N-0} & \texttt{order\_labs} & baseline incorrect; Mem0/RAG correct & Memory helps when prior completed workup changes the next action. \\
        \texttt{P000007-V11-T0033-T3-N-0} & \texttt{order\_labs} & event correct; note incorrect & Notes may over-emphasize diagnosis while events preserve monitoring needs. \\
        \bottomrule
    \end{tabular*}}
\end{table}

Taken together, these cases support the retrieval--reasoning gap discussed in Chapter~3. Memory changes the information available to the agent, but the model still has to decide whether that information is current, action-relevant, and causally connected to the next clinical step. The same mechanism can therefore improve one instance, harm another, and leave the aggregate average nearly unchanged.

